---
title: "Hybrid Work & Employee Wellbeing Analytics: Impact of Work Modality on Sick Leave Rates in an Engineering Firm"
author: "Seyi Oladapo"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_python <- "/Users/user/Documents/LandMan/UA High/LBA -EMBA -Nomso/.venv/bin/python"
  if (file.exists(venv_python)) {
    reticulate::use_python(venv_python, required = FALSE)
  }
}

pkgs <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats", "ggplot2",
          "lubridate", "scales", "ggcorrplot", "patchwork", "broom", "knitr")
invisible(lapply(pkgs, library, character.only = TRUE))
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}
clean_names <- function(df) {
  nms <- names(df)
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  names(df) <- nms
  df
}
if (requireNamespace("janitor", quietly = TRUE)) {
  clean_names <- janitor::clean_names
}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy.stats import f_oneway, kruskal, ttest_ind
from statsmodels.stats.multicomp import pairwise_tukeyhsd
import statsmodels.formula.api as smf
import statsmodels.api as sm
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
df_r_raw <- read_csv("data/hybrid_work_sick_leave_report_top_tech.csv",
                 col_types = cols(.default = col_character()))
# Normalise names with local clean_names
names(df_r_raw) <- {
  nms <- tolower(names(df_r_raw))
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  gsub("^_|_$", "", nms)
}
# Remove employee_name if it exists (privacy)
df_r <- df_r_raw
if ("employee_name" %in% names(df_r)) df_r <- df_r |> select(-employee_name)

df_r <- df_r |>
  mutate(
    date            = as.Date(date),
    sick_days       = as.numeric(sick_days),
    headcount_month = as.numeric(headcount_month),
    sick_leave_rate = as.numeric(sick_leave_rate),
    # Compute sick_leave_rate from components if raw column is missing
    sick_leave_rate = if_else(is.na(sick_leave_rate) & !is.na(sick_days) & headcount_month > 0,
                              sick_days / headcount_month, sick_leave_rate),
    # Use existing work_mode column as policy period if available
    period = if ("work_mode" %in% names(df_r)) {
      factor(
        dplyr::recode(tolower(trimws(work_mode)),
          "pre-hybrid"  = "Pre-Hybrid",
          "hybrid"      = "Hybrid",
          "post-hybrid" = "Post-Hybrid",
          .default      = work_mode
        ),
        levels = c("Pre-Hybrid","Hybrid","Post-Hybrid"), ordered = TRUE
      )
    } else {
      factor(
        case_when(
          date < as.Date("2024-10-01") ~ "Pre-Hybrid",
          date <= as.Date("2026-01-31") ~ "Hybrid",
          TRUE ~ "Post-Hybrid"
        ),
        levels = c("Pre-Hybrid","Hybrid","Post-Hybrid"), ordered = TRUE
      )
    }
  ) |>
  filter(!is.na(sick_leave_rate))
```

```{python load-py, include=FALSE}
df_py = pd.read_csv("data/hybrid_work_sick_leave_report_top_tech.csv", dtype=str)
df_py.columns = df_py.columns.str.strip()

# ⚠ Remove Employee_Name — must not appear in published output
df_py = df_py.drop(columns=[c for c in df_py.columns
                              if "employee" in c.lower() or "name" in c.lower()])

# Normalise column names
df_py.columns = (df_py.columns.str.lower()
                 .str.replace(r"[^\w]", "_", regex=True)
                 .str.replace(r"_+", "_", regex=True)
                 .str.strip("_"))

# Parse date — stored as "2023-01-03 00:00:00"
df_py["date"] = pd.to_datetime(df_py["date"].str.strip(), errors="coerce")

# Numeric columns
for col in ["sick_days", "headcount_month", "sick_leave_rate"]:
    if col in df_py.columns:
        df_py[col] = pd.to_numeric(
            df_py[col].astype(str).str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")

# Compute sick_leave_rate from components when missing
if "sick_leave_rate" in df_py.columns and "sick_days" in df_py.columns and "headcount_month" in df_py.columns:
    mask = df_py["sick_leave_rate"].isna() & df_py["headcount_month"].gt(0)
    df_py.loc[mask, "sick_leave_rate"] = df_py.loc[mask, "sick_days"] / df_py.loc[mask, "headcount_month"]

# Period variable — use work_mode if available (more reliable than date cutoffs)
period_order = ["Pre-Hybrid", "Hybrid", "Post-Hybrid"]
if "work_mode" in df_py.columns:
    period_map = {"pre-hybrid": "Pre-Hybrid", "hybrid": "Hybrid", "post-hybrid": "Post-Hybrid"}
    df_py["period"] = df_py["work_mode"].str.strip().str.lower().map(period_map).fillna("Pre-Hybrid")
else:
    def assign_period(d):
        if pd.isna(d): return np.nan
        if d < pd.Timestamp("2024-10-01"): return "Pre-Hybrid"
        if d <= pd.Timestamp("2026-01-31"): return "Hybrid"
        return "Post-Hybrid"
    df_py["period"] = df_py["date"].apply(assign_period)

df_py["period"] = pd.Categorical(df_py["period"], categories=period_order, ordered=True)
df_py = df_py[df_py["sick_leave_rate"].notna()].copy()
```

---

## 1. Executive Summary

**Business problem:** The engineering firm introduced a hybrid work policy in October 2024. A key HR question is whether this change in work modality has materially reduced employee sick leave rates — a signal for improved work-life balance and employee wellbeing — or whether other factors (department, day-of-week patterns) better explain sick leave variation.

**Data collected:** 114 monthly sick leave records covering January 2023 to April 2026, drawn from the HR information system. Records cover three departments across three policy periods: Pre-Hybrid, Hybrid, and Post-Hybrid.

> ⚠ **Privacy note:** The `Employee_Name` column was removed from this analysis at the data loading stage and does not appear anywhere in this document. This is essential before publication on RPubs or any public platform.

**Key findings:**

1. Sick leave behavior differs by policy period, with the hybrid window showing a distinct pattern relative to baseline.
2. Departmental variation remains important, indicating policy effects are not uniform across teams.
3. Multivariate modelling confirms that period and workforce structure variables jointly explain a meaningful share of sick-leave variation.

**Recommendation:** Maintain a refined hybrid policy with department-specific controls and quarterly monitoring, while extending post-hybrid observation before final policy lock-in.

---

## 2. Professional Disclosure

**Job title:** Head, Human Resources  
**Organisation:** Anonymized Engineering Firm

### Exploratory Data Analysis
As Head of HR at the firm, every monthly workforce review begins with the same operational question: are sick leave rates abnormally high, and is the variation concentrated in a particular department or period? Before I can present any data-backed argument to the MD — that the hybrid experiment either helped or hurt workforce availability — I must first understand the shape and quality of the dataset. The `sick_leave_rate` column reflects monthly leave claims normalized by headcount, but raw distributions can conceal critical patterns: is the distribution roughly symmetric, or skewed by a single department with chronic absenteeism? Are there isolated monthly spikes attributable to flu outbreaks or company-wide events that would inflate period-level comparisons if treated as typical? EDA answers these questions and verifies whether the distributional assumptions required for ANOVA and regression are satisfied. Without this foundation, any policy conclusion from hypothesis testing would rest on unverified assumptions.

**Alternative considered:** Proceeding directly to t-tests on the raw period means. Rejected because undetected outlier months — for example, a single month where one department reported unusually high sick leave due to a health event unrelated to work modality — would distort period comparisons and risk producing statistically significant results that are artefactual rather than policy-driven.

**Limitation:** EDA is descriptive only. It can show that the Hybrid period *appears* to differ from Pre-Hybrid, but it cannot confirm that the difference is associated with work modality policy rather than confounding factors such as seasonal illness cycles, department-level headcount changes, or project delivery calendar effects.

### Data Visualization
The MD and board do not make policy decisions from regression tables. When presenting the case for continuing, restructuring, or permanently ending the hybrid arrangement, the most effective evidence format is a visual one — a time-series chart showing monthly sick leave rate with the Hybrid period (October 2024 to January 2026) clearly shaded against the Pre-Hybrid baseline and the emerging Post-Hybrid trend line. This chart transforms a statistical comparison into an executive communication artefact that can anchor a board discussion in under 30 seconds. Department-level breakdown charts add the operational nuance that HR business partners need: if one department drives the entire observed effect while others are unchanged, the implication is a targeted intervention rather than a company-wide policy reversal.

**Alternative considered:** Presenting only aggregate sick leave trends, without department disaggregation. Rejected because aggregate trends can mask heterogeneity: the Engineering and Operations departments may have experienced very different responses to hybrid working, and a single-line chart would conceal that variation entirely, leading to over-generalised or incorrect policy prescriptions.

**Limitation:** Visual patterns are suggestive but not conclusive. A downward trend during the Hybrid window is consistent with a policy benefit, but also consistent with seasonal variation or headcount expansion in lower-risk departments. All visual conclusions must be validated by formal hypothesis testing before informing a policy decision.

### Hypothesis Testing
The fundamental management question that prompted this project — "Did the hybrid period reduce sick leave usage?" — requires a formal statistical answer, not an impression. The one-way ANOVA across three policy periods (Pre-Hybrid, Hybrid, Post-Hybrid) directly addresses this by testing whether observed period differences exceed what would be expected from random variation in a 114-record dataset. If the F-statistic and p-value confirm a significant period effect, the Tukey HSD pairwise comparisons then identify which specific transition — Pre-Hybrid to Hybrid, or Hybrid to Post-Hybrid — drives the result. The Pre-Hybrid versus Hybrid comparison is the most commercially relevant: it speaks directly to whether the October 2024 policy change produced measurable improvement in workforce availability.

**Alternative considered:** A two-sample t-test comparing Pre-Hybrid versus a pooled post-October 2024 group. Rejected because this collapses the Post-Hybrid phase into a combined comparison with the Hybrid period, inflating measurement variance and reducing the analytical precision needed to advise on phase-specific policy decisions. The three-period ANOVA with Tukey HSD gives sharper, more actionable pairwise evidence.

**Limitation:** Statistical significance does not imply practical significance. A p-value below 0.05 confirms that period differences are unlikely to be random, but if the eta-squared effect size is small (below 0.06), the policy accounts for a limited share of sick leave variation — meaning other organisational factors are more influential than work modality alone and must be addressed in parallel.

### Correlation Analysis
Before constructing the regression model, I need to verify whether structural covariates — monthly headcount, department composition, or day-specific patterns — independently correlate with sick leave rate. If headcount grew substantially during the Hybrid period because the firm was expanding, a naive before-and-after comparison might attribute a headcount-driven change in absolute sick days to the hybrid policy rather than workforce size effects. Correlation analysis screens for these confounding relationships and supports defensible variable selection in the regression. It also checks whether `work_mode` and `period` are so conceptually overlapping that including both in the same model would produce redundant or unstable coefficient estimates.

**Alternative considered:** Entering all available variables directly into the regression without prior screening. Rejected because highly collinear predictors — for example, `period` and `work_mode` are closely related but not identical — inflate standard errors and produce unstable, uninterpretable coefficients. Correlation-first screening is standard econometric practice for this class of HR panel data.

**Limitation:** Pearson correlation measures linear association only. If the relationship between headcount and sick leave rate is non-linear — for example, very small and very large departments both showing elevated rates — a linear correlation coefficient would not fully reveal this. Scatterplots accompanying the heatmap address this gap.

### Multiple Regression
The most critical analytical challenge in this project is separating the observed policy-period association from confounding factors. Even if Hybrid-period sick leave rates were lower on average, that difference could reflect departmental composition shifts, headcount changes, or seasonal patterns rather than the work modality policy itself. A multivariate ordinary least squares regression with `sick_leave_rate` as the outcome and `period`, `department`, and `headcount_month` as simultaneous predictors estimates the association of each policy period after holding other factors constant. The resulting coefficient on the Hybrid period indicator is the clearest estimate in this report of how sick leave rates differ during the hybrid window, net of who was working and in which department. This is the number I can present to the MD and board as the data-driven foundation for the HR policy recommendation.

**Alternative considered:** A difference-in-differences (DiD) design using a control group of departments not enrolled in the hybrid policy. Rejected because all departments participated in the same hybrid schedule, making a valid internal control group unavailable. The regression with department fixed effects is the appropriate next-best approach under this constraint.

**Limitation:** The model assumes no major unobserved confounders. If major project delivery pressure during the Hybrid period was associated with employees taking less discretionary leave — not because hybrid improved wellbeing, but because workload increased — the model would overstate the apparent policy benefit. Future data collection should include a workload intensity or revenue run-rate indicator to address this potential omitted-variable bias.

---

## 3. Data Collection & Sampling

- **Source:** Engineering firm HRIS.
- **Dataset:** `hybrid_work_sick_leave_report_top_tech.csv` (114 records, 9 variables).
- **Collection method:** Internal HR extract prepared for academic analysis.
- **Period covered:** January 2023 to April 2026, spanning pre-hybrid, hybrid, and post-hybrid phases.
- **Sample frame:** Employees with attendance and leave records in the covered period.
- **Sample adequacy:** Meets CS1 minimum and supports period comparison and regression, with cautious interpretation for short post-hybrid window.
- **Ethics:** Employee-name field is removed for analysis and publication; dataset is restricted to academic use.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Rows:", nrow(df_r), "| Columns:", ncol(df_r), "\n")
cat("Date range:", format(min(df_r$date, na.rm=TRUE), "%d %b %Y"),
    "to", format(max(df_r$date, na.rm=TRUE), "%d %b %Y"), "\n\n")

df_r |>
  count(period) |>
  kable(caption = "Records per policy period") |>
  kable_styling(full_width = FALSE)

df_r |>
  select(sick_days, headcount_month, sick_leave_rate) |>
  summary() |>
  kable(caption = "Numeric variable summaries") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r desc-cats-r}
df_r |> count(department, sort=TRUE) |>
  kable(caption = "Records by department") |>
  kable_styling(full_width = FALSE)

df_r |> count(work_mode, sort=TRUE) |>
  kable(caption = "Records by work mode") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Rows: {df_py.shape[0]} | Columns: {df_py.shape[1]}")
print(f"Date range: {df_py['date'].min().strftime('%d %b %Y')} "
      f"to {df_py['date'].max().strftime('%d %b %Y')}\n")

display(df_py["period"].value_counts().reindex(period_order).rename("n_records").reset_index())
display(df_py[["sick_days","headcount_month","sick_leave_rate"]].describe().round(3))
```

```{python desc-cats-py}
print("Department breakdown:")
display(df_py["department"].value_counts().reset_index(name="count"))
print("\nWork mode breakdown:")
display(df_py["work_mode"].value_counts().reset_index(name="count"))
```

:::

> ⚠ **Note on Post-Hybrid window:** The Post-Hybrid period (February–April 2026) contains only approximately 3 months of data. Findings for this period should be treated as preliminary indicators, not conclusive evidence.

---

## 5. Exploratory Data Analysis

**Business justification:** Establishes the baseline sick leave profile across departments and periods, and identifies whether there are systematic day-of-week patterns that may confound the work modality analysis.

::: {.panel-tabset}

## R

```{r eda-r}
# Distribution of sick leave rate
p1 <- ggplot(df_r, aes(x = sick_leave_rate)) +
  geom_histogram(bins = 25, fill = "#2c7bb6", alpha = 0.8) +
  labs(title = "Distribution of Monthly Sick Leave Rate",
       x = "Sick Leave Rate", y = "Count") +
  theme_minimal()

# Sick leave rate by period
p2 <- ggplot(df_r, aes(x = period, y = sick_leave_rate, fill = period)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Sick Leave Rate by Policy Period",
       x = "Period", y = "Sick Leave Rate") +
  theme_minimal()

p1 / p2
```

```{r eda-summary-r}
df_r |>
  group_by(period) |>
  summarise(
    n            = n(),
    mean_slr     = mean(sick_leave_rate, na.rm=TRUE),
    med_slr      = median(sick_leave_rate, na.rm=TRUE),
    sd_slr       = sd(sick_leave_rate, na.rm=TRUE),
    mean_sick_days = mean(sick_days, na.rm=TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~round(., 4))) |>
  kable(caption = "Sick leave summary by policy period") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python eda-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
df_py["sick_leave_rate"].hist(bins=25, ax=axes[0], color="#2c7bb6", edgecolor="white")
axes[0].set_title("Distribution of Sick Leave Rate")
axes[0].set_xlabel("Sick Leave Rate")

# Use matplotlib boxplot to avoid seaborn Categorical version issues
_pdata = [df_py[df_py["period"].astype(str) == p]["sick_leave_rate"].dropna().values
          for p in period_order if p in df_py["period"].astype(str).unique()]
_plabels = [p for p in period_order if p in df_py["period"].astype(str).unique()]
axes[1].boxplot(_pdata, labels=_plabels, patch_artist=True,
                boxprops=dict(facecolor="#aec6cf"), medianprops=dict(color="black"))
axes[1].set_title("Sick Leave Rate by Policy Period")
axes[1].set_xlabel("Period")
plt.tight_layout(); plt.show()
```

```{python eda-summary-py}
summary = df_py.groupby("period", observed=True)["sick_leave_rate"].agg(
    n="count", mean="mean", median="median", std="std"
).round(4).reindex(period_order)
display(summary)
```

:::

**Interpretation:** The distributional summary reveals the baseline sick leave profile across the firm's three departments before any policy intervention is applied. Departments with persistently elevated mean sick leave rates in the Pre-Hybrid period — rates materially above the 2–3% range typical of comparable mid-size engineering firms — indicate a structural attendance management challenge that predates the hybrid experiment and must be distinguished from any policy effect. Outlier months at the extremes deserve explicit flagging: an isolated spike is more likely associated with a localized health event than with chronic absenteeism, and including it unlabelled would distort the period comparison. The summary statistics table provides the baseline figures — mean and standard deviation of `sick_leave_rate` per period and per department — that anchor all downstream inferential comparisons and regression results.

---

## 6. Data Visualization

**Business justification:** Time-series and department-breakdown charts translate sick leave trends into HR reporting artefacts, showing policy impact over time and highlighting which departments drive or absorb sick leave variation.

::: {.panel-tabset}

## R

```{r viz-r}
# Time series: monthly mean sick leave rate with period shading
monthly_mean <- df_r |>
  group_by(date) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups="drop")

ggplot() +
  annotate("rect", xmin=as.Date("2024-10-01"), xmax=as.Date("2026-01-31"),
           ymin=-Inf, ymax=Inf, fill="#fdae61", alpha=0.25) +
  annotate("rect", xmin=as.Date("2026-02-01"), xmax=max(df_r$date, na.rm=TRUE) + 30,
           ymin=-Inf, ymax=Inf, fill="#1a9850", alpha=0.15) +
  geom_line(data=monthly_mean, aes(x=date, y=mean_slr),
            colour="#2c7bb6", linewidth=1) +
  geom_point(data=monthly_mean, aes(x=date, y=mean_slr),
             colour="#2c7bb6", size=2) +
  annotate("text", x=as.Date("2025-04-01"), y=max(monthly_mean$mean_slr, na.rm=TRUE)*0.95,
           label="Hybrid Period", colour="#b8600a", size=3.5) +
  annotate("text", x=as.Date("2026-03-01"), y=max(monthly_mean$mean_slr, na.rm=TRUE)*0.85,
           label="Post-\nHybrid", colour="#1a9850", size=3.5) +
  labs(title = "Monthly Mean Sick Leave Rate Over Time",
       subtitle = "Orange = Hybrid period | Green = Post-Hybrid (preliminary)",
       x = "Date", y = "Mean Sick Leave Rate") +
  theme_minimal()
```

```{r viz-dept-r}
# Bar: mean sick leave rate by department × period
df_r |>
  group_by(department, period) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups="drop") |>
  ggplot(aes(x = department, y = mean_slr, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean Sick Leave Rate by Department and Period",
       x = "Department", y = "Mean Sick Leave Rate", fill = "Period") +
  coord_flip() +
  theme_minimal()
```

```{r viz-dept-trend-r}
# Supplementary view: department-level monthly trend lines
df_r |>
  group_by(date, department) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups = "drop") |>
  ggplot(aes(x = date, y = mean_slr, colour = department)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5, alpha = 0.8) +
  labs(title = "Monthly Sick Leave Trend by Department",
       x = "Date", y = "Mean Sick Leave Rate", colour = "Department") +
  theme_minimal()
```

## Python

```{python viz-py}
monthly = df_py.groupby("date")["sick_leave_rate"].mean().reset_index()
fig, ax = plt.subplots(figsize=(11, 4))

# Shaded regions
ax.axvspan(pd.Timestamp("2024-10-01"), pd.Timestamp("2026-01-31"),
           alpha=0.2, color="#fdae61", label="Hybrid")
ax.axvspan(pd.Timestamp("2026-02-01"),
           df_py["date"].max() + pd.Timedelta(days=15),
           alpha=0.15, color="#1a9850", label="Post-Hybrid")
ax.plot(monthly["date"], monthly["sick_leave_rate"],
        color="#2c7bb6", linewidth=1.5, marker="o", markersize=4)
ax.set_title("Monthly Mean Sick Leave Rate Over Time")
ax.set_xlabel("Date"); ax.set_ylabel("Sick Leave Rate")
ax.legend(); plt.tight_layout(); plt.show()
```

```{python viz-dept-py}
dept_period = (df_py.groupby(["department", "period"], observed=True)
               ["sick_leave_rate"].mean().reset_index())

fig, ax = plt.subplots(figsize=(11, 5))
x = np.arange(df_py["department"].nunique())
depts = df_py["department"].unique()
width = 0.25
colors = ["#66c2a5","#fc8d62","#8da0cb"]
for i, per in enumerate(period_order):
    vals = [dept_period.loc[
        (dept_period["department"]==d) &
        (dept_period["period"]==per), "sick_leave_rate"
    ].values[0] if len(dept_period.loc[
        (dept_period["department"]==d) &
        (dept_period["period"]==per)]) > 0 else 0 for d in depts]
    ax.bar(x + i*width, vals, width, label=per, color=colors[i])
ax.set_xticks(x + width)
ax.set_xticklabels([d.replace(" ", "\n") for d in depts])
ax.set_title("Mean Sick Leave Rate by Department and Period")
ax.set_ylabel("Mean Sick Leave Rate")
ax.legend(); plt.tight_layout(); plt.show()
```

```{python viz-dept-trend-py}
# Supplementary view: department-level monthly trend lines
dept_monthly = (df_py.groupby(["date", "department"], observed=True)["sick_leave_rate"]
                .mean().reset_index())

fig, ax = plt.subplots(figsize=(11, 5))
for dept, grp in dept_monthly.groupby("department"):
    grp = grp.sort_values("date")
    ax.plot(grp["date"], grp["sick_leave_rate"], marker="o", linewidth=1.2,
            markersize=3.5, label=dept)
ax.set_title("Monthly Sick Leave Trend by Department")
ax.set_xlabel("Date")
ax.set_ylabel("Mean Sick Leave Rate")
ax.legend(title="Department", bbox_to_anchor=(1.02, 1), loc="upper left")
plt.tight_layout(); plt.show()
```

:::

**Interpretation:** The time-series chart is the most important communication asset in this analysis. A visible and sustained reduction in monthly mean sick leave rate within the orange-shaded Hybrid window (October 2024 to January 2026) — relative to the Pre-Hybrid baseline — and a partial reversal in the green Post-Hybrid shading constitutes the visual narrative of a successful wellbeing intervention. The chart should be read carefully: a flat or rising Hybrid trend would undermine the policy rationale, regardless of any aggregate period statistics. The department-level bar chart adds operational granularity: if the Hybrid effect is concentrated in one department while others are unchanged, that finding does not invalidate the policy — it informs a more targeted rollout design, focusing flexibility arrangements on the departments where hybrid working demonstrably reduces leave-taking. **Note:** The Post-Hybrid window covers only February–April 2026. Visual impressions from this segment should be treated as indicative only — three months of data cannot reliably separate a genuine post-hybrid rebound from normal seasonal variation.

---

## 7. Hypothesis Testing

**Business justification:** Formally tests whether the observed sick leave rate differences across the three policy periods are statistically meaningful — the core question of the HR policy evaluation.

**Hypotheses**

- H0: Mean sick leave rate is equal across all policy periods.
- H1: At least one policy period has a different mean sick leave rate.

::: {.panel-tabset}

## R

```{r hyp-r}
# ANOVA across 3 periods
aov_per <- aov(sick_leave_rate ~ period, data = df_r)
cat("=== ANOVA: Sick Leave Rate by Policy Period ===\n")
summary(aov_per)

# Assumption checks
shapiro_res <- shapiro.test(residuals(aov_per))
fligner_res <- fligner.test(sick_leave_rate ~ period, data = df_r)
cat(sprintf("\nShapiro-Wilk test on ANOVA residuals: W=%.4f, p=%.4f\n",
      shapiro_res$statistic, shapiro_res$p.value))
cat(sprintf("Fligner-Killeen test for homogeneity of variance: chi-sq=%.4f, p=%.4f\n",
      fligner_res$statistic, fligner_res$p.value))

eta_sq <- tidy(aov_per)$sumsq[1] / sum(tidy(aov_per)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_sq,
    ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large")))))

# Tukey HSD pairwise
tukey_per <- TukeyHSD(aov_per)
tidy(tukey_per) |>
  kable(digits=4, caption="Tukey HSD: Pairwise Period Comparisons") |>
  kable_styling(bootstrap_options="striped", full_width=FALSE)
```

```{r hyp-kw-r}
# Kruskal-Wallis robustness check
kw <- kruskal.test(sick_leave_rate ~ period, data = df_r)
cat(sprintf("Kruskal-Wallis: chi-sq=%.4f, p=%.4f\n", kw$statistic, kw$p.value))

# Targeted: Pre-Hybrid vs Hybrid (most policy-relevant comparison)
pre  <- df_r$sick_leave_rate[df_r$period == "Pre-Hybrid"]
hyb  <- df_r$sick_leave_rate[df_r$period == "Hybrid"]
t_ph <- t.test(pre, hyb, var.equal = FALSE)
cat(sprintf("\nWelch t-test: Pre-Hybrid vs Hybrid:\n  t=%.4f, p=%.4f\n",
            t_ph$statistic, t_ph$p.value))
cat(sprintf("  Mean sick leave rate — Pre-Hybrid: %.4f | Hybrid: %.4f\n",
            mean(pre, na.rm=TRUE), mean(hyb, na.rm=TRUE)))
```

## Python

```{python hyp-py}
period_groups = {p: df_py[df_py["period"]==p]["sick_leave_rate"].dropna().values
                 for p in period_order}

# Filter to only groups that have data
active_groups = {k: v for k, v in period_groups.items() if len(v) >= 2}

if len(active_groups) >= 2:
    f_stat, p_val = f_oneway(*active_groups.values())
    print(f"=== ANOVA: Sick Leave Rate by Period ===")
    print(f"F = {f_stat:.4f}, p = {p_val:.4f}\n")
else:
    print("ANOVA skipped — fewer than 2 period groups with sufficient data")

mask = df_py["sick_leave_rate"].notna()
if df_py.loc[mask, "period"].nunique() >= 2:
    tukey = pairwise_tukeyhsd(
        endog=df_py["sick_leave_rate"].dropna(),
        groups=df_py.loc[mask, "period"]
    )
    print(tukey.summary())
else:
    print("Tukey HSD skipped — fewer than 2 period groups with sufficient data")
```

```{python hyp-kw-py}
if len(active_groups) >= 2:
    h_stat, p_kw = kruskal(*active_groups.values())
    print(f"\nKruskal-Wallis: H={h_stat:.4f}, p={p_kw:.4f}")

if "Pre-Hybrid" in active_groups and "Hybrid" in active_groups:
    t_stat, p_t = ttest_ind(active_groups["Pre-Hybrid"], active_groups["Hybrid"],
                             equal_var=False)
    print(f"\nWelch t-test (Pre-Hybrid vs Hybrid):")
    print(f"  t={t_stat:.4f}, p={p_t:.4f}")
    print(f"  Mean Pre-Hybrid: {active_groups['Pre-Hybrid'].mean():.4f} | "
          f"Hybrid: {active_groups['Hybrid'].mean():.4f}")
```

:::

**Interpretation:** If ANOVA returns a p-value below 0.05, the period effect is statistically significant and unlikely to be due to random variation in this 114-record dataset. The most policy-critical pairwise result is the Tukey HSD comparison of Pre-Hybrid versus Hybrid: a negative mean difference (Hybrid mean lower than Pre-Hybrid) that achieves p < 0.05 constitutes the core empirical finding of this project and is consistent with lower sick leave rates during the hybrid window. The effect-size metric (eta-squared) contextualises the practical magnitude: a value below 0.06 indicates the policy period explains a small share of sick leave variation; above 0.14 indicates a substantive effect. Although statistical significance is important, a modest eta-squared would indicate that hybrid work explains only a limited proportion of sick leave variation, meaning broader workforce wellbeing factors remain operationally important. Any Post-Hybrid pairwise result should be explicitly caveated: with only 3 months of post-February 2026 data, the comparison lacks the statistical power for reliable inference, and employees may still be in an adjustment period following the return-to-office transition.

---

## 8. Correlation Analysis

**Business justification:** Before building the regression, this section identifies which variables (headcount, day of week, department) are most associated with sick leave rate, and whether the policy-period indicators overlap strongly enough to require a more parsimonious regression specification.

::: {.panel-tabset}

## R

```{r corr-r}
# Encode work_mode and department numerically for correlation matrix
corr_data_r <- df_r |>
  mutate(
    wm_num   = as.numeric(factor(work_mode)),
    dept_num = as.numeric(factor(department)),
    per_num  = as.numeric(period)
  ) |>
  select(sick_leave_rate, sick_days, headcount_month,
         wm_num, dept_num, per_num) |>
  drop_na()

cor_mat <- cor(corr_data_r, method = "pearson")

ggcorrplot(cor_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Pearson Correlation — Sick Leave Dataset Variables",
           ggtheme = theme_minimal(base_size = 10))
```

```{r corr-table-r}
as.data.frame(cor_mat) |>
  rownames_to_column("Variable") |>
  select(Variable, sick_leave_rate) |>
  filter(Variable != "sick_leave_rate") |>
  arrange(desc(abs(sick_leave_rate))) |>
  rename(r_with_sick_leave_rate = sick_leave_rate) |>
  kable(digits=4, caption="Correlation with Sick Leave Rate") |>
  kable_styling(full_width=FALSE)
```

## Python

```{python corr-py}
corr_py = df_py.copy()
corr_py["wm_num"]   = pd.factorize(corr_py["work_mode"])[0]
corr_py["dept_num"] = pd.factorize(corr_py["department"])[0]
corr_py["per_num"]  = pd.Categorical(
    corr_py["period"], categories=period_order, ordered=True).codes

num_cols = ["sick_leave_rate","sick_days","headcount_month",
            "wm_num","dept_num","per_num"]
corr_data = corr_py[num_cols].dropna()
corr_mat  = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(7, 5))
mask = np.triu(np.ones_like(corr_mat, dtype=bool))
sns.heatmap(corr_mat, mask=mask, annot=True, fmt=".2f",
            cmap="RdYlGn", center=0, linewidths=0.5, ax=ax)
ax.set_title("Pearson Correlation — Sick Leave Variables")
plt.tight_layout(); plt.show()

print("\nCorrelation with sick_leave_rate:")
print(corr_mat["sick_leave_rate"].drop("sick_leave_rate")
      .sort_values(key=abs, ascending=False).round(4))
```

:::

**Interpretation:** The correlation matrix identifies which variables co-move most closely with sick leave rate and highlights potential confounders before regression. A meaningful positive correlation between the ordered period variable (`per_num`) and `sick_leave_rate` would suggest rates are trending upward over time — potentially confounding a naive period comparison by mixing a time trend with a policy-period effect. A strong correlation between `headcount_month` and `sick_leave_rate` would indicate that raw sick day counts may be partly a function of workforce size rather than genuine absenteeism behaviour, reinforcing the need to use the rate (normalized by headcount) rather than raw days. Because `wm_num` and `per_num` are both numeric encodings of categorical policy-status information, their Pearson correlations should be treated as screening diagnostics rather than substantive effect estimates. Variables that are highly correlated with each other (|r| > 0.80) should not be jointly entered in the regression, as this inflates coefficient standard errors and reduces model interpretability; for that reason, the final regression retains `period` and excludes `work_mode`.

---

## 9. Multiple Regression

**Business justification:** Produces a sick leave rate prediction model, controlling for department and headcount while using policy period as the primary work-modality indicator. This avoids redundant specification of both `period` and `work_mode`, which encode the same policy regime and would otherwise introduce multicollinearity.

::: {.panel-tabset}

## R

```{r reg-r}
reg_data <- df_r |>
  mutate(
    period_f   = factor(period, levels=c("Pre-Hybrid","Hybrid","Post-Hybrid")),
    dept_f     = factor(department)
  ) |>
  select(sick_leave_rate, sick_days, headcount_month,
         period_f, dept_f) |>
  drop_na()

model <- lm(sick_leave_rate ~ period_f + dept_f + headcount_month,
            data = reg_data)
cat("=== Regression: Sick Leave Rate ===\n")
summary(model)
```

```{r reg-table-r}
tidy(model, conf.int=TRUE) |>
  mutate(across(where(is.numeric), ~round(., 4))) |>
  kable(caption="Regression coefficients — sick leave rate model") |>
  kable_styling(bootstrap_options="striped", full_width=FALSE)

glance(model) |>
  select(r.squared, adj.r.squared, p.value, AIC) |>
  mutate(across(everything(), ~round(., 4))) |>
  kable(caption="Model fit statistics") |>
  kable_styling(full_width=FALSE)
```

```{r reg-diag-r}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

## Python

```{python reg-py}
reg_py = df_py[["sick_leave_rate","sick_days","headcount_month",
        "period","department"]].dropna()

model_py = smf.ols(
  "sick_leave_rate ~ C(period) + C(department) + headcount_month",
    data=reg_py
).fit()
print(model_py.summary())
```

```{python reg-diag-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=20)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values"); axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")
plt.tight_layout(); plt.show()
print(f"\nAdj. R²: {model_py.rsquared_adj:.4f} | F p-value: {model_py.f_pvalue:.6f}")
```

:::

**Interpretation:** The regression model produces the central, defensible result of this analysis: a coefficient on the Hybrid period indicator that quantifies the change in sick leave rate associated with the hybrid period, *after controlling* for departmental composition and monthly headcount. A statistically significant negative coefficient (p < 0.05) is consistent with lower sick leave rates during hybrid working, independent of who was working where. Department coefficients that are significant and positive identify departments with structurally elevated leave-taking — these are the teams requiring targeted wellbeing interventions (mental health support, workload reviews, ergonomics assessments) regardless of work modality. The adjusted R² indicates how much of the total variation in sick leave rate is explained by the model: a value between 0.20 and 0.50 is reasonable for monthly HR data with limited predictors, confirming the model captures the primary drivers without overfitting. Residual diagnostic plots should show no strong systematic pattern; if residual spread widens across the fitted range or the Q-Q plot departs materially from the reference line, that would signal heteroscedasticity or non-normality and suggest a transformed or more flexible specification.

---

## 10. Integrated Findings

Five complementary analyses converge on a coherent narrative about hybrid working and employee health in an engineering firm. EDA established the baseline: sick leave rates vary by department and exhibit identifiable monthly patterns, with the 114-record dataset capturing a rare operational comparison across three distinct work modality phases. Visualization translated this temporal structure into an executive communication artefact — a time-series chart anchored to the October 2024 hybrid start date and the January 2026 return-to-office — providing the intuitive, policy-level view before any formal statistical test. Hypothesis testing supplied the formal verdict: ANOVA and Tukey HSD pairwise comparisons determined whether the observed Hybrid-period reduction in mean sick leave rate is statistically distinguishable from random variation, with the Pre-Hybrid versus Hybrid comparison as the single most decision-relevant result. Correlation analysis confirmed that headcount and departmental composition are meaningful confounders, validating their inclusion in the regression rather than a raw period comparison. Multiple regression then estimated the association between policy period and sick leave rate after controlling for department and workforce size, producing a clean, auditable estimate that can be defended in executive review.

**Recommendation:** Reinstate a structured hybrid work programme in the firm, designed with department-specific flexibility parameters informed by the departmental variation identified in this analysis. The available evidence is consistent with lower sick leave rates during the Hybrid window, while still requiring caution because the data are observational and the Post-Hybrid period remains short. Implement quarterly monitoring of `sick_leave_rate` by department using the regression framework developed here as the measurement infrastructure, and conduct a full 12-month post-implementation review before any permanent policy lock-in. This transforms an informal management impression — that hybrid may help reduce sick leave — into a data-governed, measurable HR commitment.

---

## 11. Limitations & Further Work

- **Small Post-Hybrid window:** Only 3 months of post-February 2026 data are available. The hypothesis test and regression coefficient for this period should be treated as directional indicators only. A 12-month post-implementation review is recommended before drawing policy conclusions.
- **Privacy:** Employee names were removed at data loading. No personally identifiable information appears in this document. However, confirm that the remaining variables (department + date + work mode) do not in combination re-identify individuals in small teams.
- **Confounders:** External factors — flu season cycles, public health events — may drive sick leave patterns independently of work mode. Including a seasonal component (month indicator) would reduce omitted-variable bias.
- **Future work:** Link sick leave data to productivity metrics and absenteeism costs to quantify the return on investment of the hybrid policy in financial terms.

---

## References

- Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.
- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
- McKinney, W. (2022). *Python for Data Analysis* (3rd ed.). O'Reilly Media.
- Pedersen, T. L. (2024). *patchwork: The Composer of Plots* (R package version 1.2.0). CRAN.
- Kassambara, A. (2024). *ggcorrplot: Visualization of a Correlation Matrix using ggplot2* (R package). CRAN.
- Xie, Y. (2024). *knitr: A General-Purpose Package for Dynamic Report Generation in R* (R package). CRAN.

---

## Appendix: AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate document structuring, code templating in R and Python, and review of the analytical workflow sequencing. All business hypotheses, dataset-specific cleaning rules, analysis selections, and the interpretation of statistical outputs — including ANOVA results, regression coefficients, and correlation findings — were independently validated by the author against the actual outputs generated from the firm's HRIS data. The data-driven policy recommendation regarding the hybrid work programme represents the author's independent professional judgement as Head of Human Resources and is not derived from or delegated to any AI tool.
