---
title: "Hybrid Work & Employee Wellbeing Analytics: Impact of Work Modality on Sick Leave Rates in an Engineering Firm"
author: "Seyi Oladapo"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_python <- ".venv/bin/python"
  if (file.exists(venv_python)) {
    reticulate::use_python(venv_python, required = FALSE)
  }
}

pkgs <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats", "ggplot2",
          "lubridate", "scales", "ggcorrplot", "patchwork", "broom", "knitr", "e1071")
invisible(lapply(pkgs, library, character.only = TRUE))
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}
clean_names <- function(df) {
  nms <- names(df)
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  names(df) <- nms
  df
}
if (requireNamespace("janitor", quietly = TRUE)) {
  clean_names <- janitor::clean_names
}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy.stats import f_oneway, kruskal, ttest_ind, levene, spearmanr, skew, shapiro
from statsmodels.stats.multicomp import pairwise_tukeyhsd
import statsmodels.formula.api as smf
import statsmodels.api as sm
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
df_r_raw <- read_csv("data/hybrid_work_sick_leave_report_top_tech.csv",
                 col_types = cols(.default = col_character()))
# Normalise names with local clean_names
names(df_r_raw) <- {
  nms <- tolower(names(df_r_raw))
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  gsub("^_|_$", "", nms)
}
# Remove employee_name if it exists (privacy)
df_r <- df_r_raw
if ("employee_name" %in% names(df_r)) df_r <- df_r |> select(-employee_name)

df_r <- df_r |>
  mutate(
    date            = as.Date(date),
    sick_days       = as.numeric(sick_days),
    headcount_month = as.numeric(headcount_month),
    sick_leave_rate = as.numeric(sick_leave_rate),
    # Compute sick_leave_rate from components if raw column is missing
    sick_leave_rate = if_else(is.na(sick_leave_rate) & !is.na(sick_days) & headcount_month > 0,
                              sick_days / headcount_month, sick_leave_rate),
    # Use existing work_mode column as policy period if available
    period = if ("work_mode" %in% names(df_r)) {
      factor(
        dplyr::recode(tolower(trimws(work_mode)),
          "pre-hybrid"  = "Pre-Hybrid",
          "hybrid"      = "Hybrid",
          "post-hybrid" = "Post-Hybrid",
          .default      = work_mode
        ),
        levels = c("Pre-Hybrid","Hybrid","Post-Hybrid"), ordered = TRUE
      )
    } else {
      factor(
        case_when(
          date < as.Date("2024-10-01") ~ "Pre-Hybrid",
          date <= as.Date("2026-01-31") ~ "Hybrid",
          TRUE ~ "Post-Hybrid"
        ),
        levels = c("Pre-Hybrid","Hybrid","Post-Hybrid"), ordered = TRUE
      )
    }
  ) |>
  filter(!is.na(sick_leave_rate))
```

```{python load-py, include=FALSE}
df_py = pd.read_csv("data/hybrid_work_sick_leave_report_top_tech.csv", dtype=str)
df_py.columns = df_py.columns.str.strip()

# ⚠ Remove Employee_Name — must not appear in published output
df_py = df_py.drop(columns=[c for c in df_py.columns
                              if "employee" in c.lower() or "name" in c.lower()])

# Normalise column names
df_py.columns = (df_py.columns.str.lower()
                 .str.replace(r"[^\w]", "_", regex=True)
                 .str.replace(r"_+", "_", regex=True)
                 .str.strip("_"))

# Parse date — stored as "2023-01-03 00:00:00"
df_py["date"] = pd.to_datetime(df_py["date"].str.strip(), errors="coerce")

# Numeric columns
for col in ["sick_days", "headcount_month", "sick_leave_rate"]:
    if col in df_py.columns:
        df_py[col] = pd.to_numeric(
            df_py[col].astype(str).str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")

# Compute sick_leave_rate from components when missing
if "sick_leave_rate" in df_py.columns and "sick_days" in df_py.columns and "headcount_month" in df_py.columns:
    mask = df_py["sick_leave_rate"].isna() & df_py["headcount_month"].gt(0)
    df_py.loc[mask, "sick_leave_rate"] = df_py.loc[mask, "sick_days"] / df_py.loc[mask, "headcount_month"]

# Period variable — use work_mode if available (more reliable than date cutoffs)
period_order = ["Pre-Hybrid", "Hybrid", "Post-Hybrid"]
if "work_mode" in df_py.columns:
    period_map = {"pre-hybrid": "Pre-Hybrid", "hybrid": "Hybrid", "post-hybrid": "Post-Hybrid"}
    df_py["period"] = df_py["work_mode"].str.strip().str.lower().map(period_map).fillna("Pre-Hybrid")
else:
    def assign_period(d):
        if pd.isna(d): return np.nan
        if d < pd.Timestamp("2024-10-01"): return "Pre-Hybrid"
        if d <= pd.Timestamp("2026-01-31"): return "Hybrid"
        return "Post-Hybrid"
    df_py["period"] = df_py["date"].apply(assign_period)

df_py["period"] = pd.Categorical(df_py["period"], categories=period_order, ordered=True)
df_py = df_py[df_py["sick_leave_rate"].notna()].copy()
```

---

## 1. Executive Summary

**Business problem:** The engineering firm introduced a hybrid work policy in October 2024. A key HR question is whether this change in work modality has materially reduced employee sick leave rates — a signal for improved work-life balance and employee wellbeing — or whether other factors (department, day-of-week patterns) better explain sick leave variation.

**Data collected:** 114 monthly sick leave records covering January 2023 to April 2026, drawn from the HR information system. Records cover 18 departments across three policy periods: Pre-Hybrid, Hybrid, and Post-Hybrid.

> ⚠ **Privacy note:** The `Employee_Name` column was removed from this analysis at the data loading stage and does not appear anywhere in this document. This is essential before publication on RPubs or any public platform.

**Key findings:**

1. Sick leave behavior differs by policy period, with the hybrid window showing a distinct pattern relative to baseline.
2. Departmental variation remains important, indicating policy effects are not uniform across teams.
3. Multivariate modelling confirms that period and workforce structure variables jointly explain a meaningful share of sick-leave variation.

**Recommendation:** Maintain a refined hybrid policy with department-specific controls and quarterly monitoring, while extending post-hybrid observation before final policy lock-in.

---

## 2. Professional Disclosure

**Job title:** Head, Human Resources  
**Organisation:** Anonymized Engineering Firm

### Exploratory Data Analysis
As Head of HR at the firm, every monthly workforce review begins with the same operational question: are sick leave rates abnormally high, and is the variation concentrated in a particular department or period? Before I can present any data-backed argument to the MD — that the hybrid experiment either helped or hurt workforce availability — I must first understand the shape and quality of the dataset. The `sick_leave_rate` column reflects monthly leave claims normalized by headcount, but raw distributions can conceal critical patterns: is the distribution roughly symmetric, or skewed by a single department with chronic absenteeism? Are there isolated monthly spikes attributable to flu outbreaks or company-wide events that would inflate period-level comparisons if treated as typical? EDA answers these questions and verifies whether the distributional assumptions required for ANOVA and regression are satisfied. Without this foundation, any policy conclusion from hypothesis testing would rest on unverified assumptions.

**Alternative considered:** Proceeding directly to t-tests on the raw period means. Rejected because undetected outlier months — for example, a single month where one department reported unusually high sick leave due to a health event unrelated to work modality — would distort period comparisons and risk producing statistically significant results that are artefactual rather than policy-driven.

**Limitation:** EDA is descriptive only. It can show that the Hybrid period *appears* to differ from Pre-Hybrid, but it cannot confirm that the difference is associated with work modality policy rather than confounding factors such as seasonal illness cycles, department-level headcount changes, or project delivery calendar effects.

### Data Visualization
The MD and board do not make policy decisions from regression tables. When presenting the case for continuing, restructuring, or permanently ending the hybrid arrangement, the most effective evidence format is a visual one — a time-series chart showing monthly sick leave rate with the Hybrid period (October 2024 to January 2026) clearly shaded against the Pre-Hybrid baseline and the emerging Post-Hybrid trend line. This chart transforms a statistical comparison into an executive communication artefact that can anchor a board discussion in under 30 seconds. Department-level breakdown charts add the operational nuance that HR business partners need: if one department drives the entire observed effect while others are unchanged, the implication is a targeted intervention rather than a company-wide policy reversal.

**Alternative considered:** Presenting only aggregate sick leave trends, without department disaggregation. Rejected because aggregate trends can mask heterogeneity: the Engineering and Operations departments may have experienced very different responses to hybrid working, and a single-line chart would conceal that variation entirely, leading to over-generalised or incorrect policy prescriptions.

**Limitation:** Visual patterns are suggestive but not conclusive. A downward trend during the Hybrid window is consistent with a policy benefit, but also consistent with seasonal variation or headcount expansion in lower-risk departments. All visual conclusions must be validated by formal hypothesis testing before informing a policy decision.

### Hypothesis Testing
The fundamental management question that prompted this project — "Did the hybrid period reduce sick leave usage?" — requires a formal statistical answer, not an impression. The one-way ANOVA across three policy periods (Pre-Hybrid, Hybrid, Post-Hybrid) directly addresses this by testing whether observed period differences exceed what would be expected from random variation in a 114-record dataset. If the F-statistic and p-value confirm a significant period effect, the Tukey HSD pairwise comparisons then identify which specific transition — Pre-Hybrid to Hybrid, or Hybrid to Post-Hybrid — drives the result. The Pre-Hybrid versus Hybrid comparison is the most commercially relevant: it speaks directly to whether the October 2024 policy change produced measurable improvement in workforce availability.

**Alternative considered:** A two-sample t-test comparing Pre-Hybrid versus a pooled post-October 2024 group. Rejected because this collapses the Post-Hybrid phase into a combined comparison with the Hybrid period, inflating measurement variance and reducing the analytical precision needed to advise on phase-specific policy decisions. The three-period ANOVA with Tukey HSD gives sharper, more actionable pairwise evidence.

**Limitation:** Statistical significance does not imply practical significance. A p-value below 0.05 confirms that period differences are unlikely to be random, but if the eta-squared effect size is small (below 0.06), the policy accounts for a limited share of sick leave variation — meaning other organisational factors are more influential than work modality alone and must be addressed in parallel.

### Correlation Analysis
Before constructing the regression model, I need to verify whether structural covariates — monthly headcount, department composition, or day-specific patterns — independently correlate with sick leave rate. If headcount grew substantially during the Hybrid period because the firm was expanding, a naive before-and-after comparison might attribute a headcount-driven change in absolute sick days to the hybrid policy rather than workforce size effects. Correlation analysis screens for these confounding relationships and supports defensible variable selection in the regression. It also checks whether `work_mode` and `period` are so conceptually overlapping that including both in the same model would produce redundant or unstable coefficient estimates.

**Alternative considered:** Entering all available variables directly into the regression without prior screening. Rejected because highly collinear predictors — for example, `period` and `work_mode` are closely related but not identical — inflate standard errors and produce unstable, uninterpretable coefficients. Correlation-first screening is standard econometric practice for this class of HR panel data.

**Limitation:** Pearson correlation measures linear association only. If the relationship between headcount and sick leave rate is non-linear — for example, very small and very large departments both showing elevated rates — a linear correlation coefficient would not fully reveal this. Scatterplots accompanying the heatmap address this gap.

### Multiple Regression
The most critical analytical challenge in this project is separating the observed policy-period association from confounding factors. Even if Hybrid-period sick leave rates were lower on average, that difference could reflect departmental composition shifts, headcount changes, or seasonal patterns rather than the work modality policy itself. A multivariate ordinary least squares regression with `sick_leave_rate` as the outcome and `period`, `department`, and `headcount_month` as simultaneous predictors estimates the association of each policy period after holding other factors constant. The resulting coefficient on the Hybrid period indicator is the clearest estimate in this report of how sick leave rates differ during the hybrid window, net of who was working and in which department. This is the number I can present to the MD and board as the data-driven foundation for the HR policy recommendation.

**Alternative considered:** A difference-in-differences (DiD) design using a control group of departments not enrolled in the hybrid policy. Rejected because all departments participated in the same hybrid schedule, making a valid internal control group unavailable. The regression with department fixed effects is the appropriate next-best approach under this constraint.

**Limitation:** The model assumes no major unobserved confounders. If major project delivery pressure during the Hybrid period was associated with employees taking less discretionary leave — not because hybrid improved wellbeing, but because workload increased — the model would overstate the apparent policy benefit. Future data collection should include a workload intensity or revenue run-rate indicator to address this potential omitted-variable bias.

---

## 3. Data Collection & Sampling

- **Source:** Engineering firm HRIS.
- **Dataset:** `hybrid_work_sick_leave_report_top_tech.csv` (114 records, 9 variables).
- **Collection method:** Internal HR extract prepared for academic analysis.
- **Period covered:** January 2023 to April 2026, spanning pre-hybrid, hybrid, and post-hybrid phases.
- **Sample frame:** Employees with attendance and leave records in the covered period.
- **Sample adequacy:** Meets CS1 minimum and supports period comparison and regression, with cautious interpretation for short post-hybrid window.
- **Ethics:** Employee-name field is removed for analysis and publication; dataset is restricted to academic use.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Rows:", nrow(df_r), "| Columns:", ncol(df_r), "\n")
cat("Date range:", format(min(df_r$date, na.rm=TRUE), "%d %b %Y"),
    "to", format(max(df_r$date, na.rm=TRUE), "%d %b %Y"), "\n\n")

df_r |>
  count(period) |>
  kable(caption = "Records per policy period") |>
  kable_styling(full_width = FALSE)

df_r |>
  select(sick_days, headcount_month, sick_leave_rate) |>
  summary() |>
  kable(caption = "Numeric variable summaries") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r desc-cats-r}
df_r |> count(department, sort=TRUE) |>
  kable(caption = "Records by department") |>
  kable_styling(full_width = FALSE)

df_r |> count(work_mode, sort=TRUE) |>
  kable(caption = "Records by work mode") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Rows: {df_py.shape[0]} | Columns: {df_py.shape[1]}")
print(f"Date range: {df_py['date'].min().strftime('%d %b %Y')} "
      f"to {df_py['date'].max().strftime('%d %b %Y')}\n")

display(df_py["period"].value_counts().reindex(period_order).rename("n_records").reset_index())
display(df_py[["sick_days","headcount_month","sick_leave_rate"]].describe().round(3))
```

```{python desc-cats-py}
print("Department breakdown:")
display(df_py["department"].value_counts().reset_index(name="count"))
print("\nWork mode breakdown:")
display(df_py["work_mode"].value_counts().reset_index(name="count"))
```

:::

> ⚠ **Note on Post-Hybrid window:** The Post-Hybrid period (February–April 2026) contains only approximately 3 months of data. Findings for this period should be treated as preliminary indicators, not conclusive evidence.

---

## 5. Exploratory Data Analysis

**Business justification:** Establishes the baseline sick leave profile across departments and periods, and identifies whether there are systematic day-of-week patterns that may confound the work modality analysis.

::: {.panel-tabset}

## R

```{r eda-r}
# Distribution of sick leave rate
p1 <- ggplot(df_r, aes(x = sick_leave_rate)) +
  geom_histogram(bins = 25, fill = "#2c7bb6", alpha = 0.8) +
  labs(title = "Distribution of Monthly Sick Leave Rate",
       x = "Sick Leave Rate", y = "Count") +
  theme_minimal()

# Sick leave rate by period
p2 <- ggplot(df_r, aes(x = period, y = sick_leave_rate, fill = period)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Sick Leave Rate by Policy Period",
       x = "Period", y = "Sick Leave Rate") +
  theme_minimal()

p1 / p2
```

```{r eda-summary-r}
df_r |>
  group_by(period) |>
  summarise(
    n            = n(),
    mean_slr     = mean(sick_leave_rate, na.rm=TRUE),
    med_slr      = median(sick_leave_rate, na.rm=TRUE),
    sd_slr       = sd(sick_leave_rate, na.rm=TRUE),
    mean_sick_days = mean(sick_days, na.rm=TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~round(., 4))) |>
  kable(caption = "Sick leave summary by policy period") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r eda-quality-r}
# ── Data Quality Issue 1: Missing values ──────────────────────────────────────
cat("=== Data Quality Issue 1 — Missing Values ===\n")
df_r |>
  summarise(across(everything(), ~ round(100 * mean(is.na(.)), 2))) |>
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") |>
  arrange(desc(pct_missing)) |>
  kable(caption = "Missing data % by column") |>
  kable_styling(full_width = FALSE)

# ── Data Quality Issue 2: IQR Outliers on sick_leave_rate ────────────────────
cat("\n=== Data Quality Issue 2 — IQR Outlier Detection (sick_leave_rate) ===\n")
q1_slr  <- quantile(df_r$sick_leave_rate, 0.25, na.rm = TRUE)
q3_slr  <- quantile(df_r$sick_leave_rate, 0.75, na.rm = TRUE)
iqr_slr <- q3_slr - q1_slr
lo_slr  <- q1_slr - 1.5 * iqr_slr
hi_slr  <- q3_slr + 1.5 * iqr_slr
n_out   <- sum(df_r$sick_leave_rate < lo_slr | df_r$sick_leave_rate > hi_slr, na.rm = TRUE)
cat(sprintf("Q1 = %.4f | Q3 = %.4f | IQR = %.4f\n", q1_slr, q3_slr, iqr_slr))
cat(sprintf("Outlier fence: [%.4f, %.4f]\n", lo_slr, hi_slr))
cat(sprintf("Outlier records: %d (%.1f%% of dataset)\n\n", n_out, 100 * n_out / nrow(df_r)))

df_r |>
  filter(sick_leave_rate < lo_slr | sick_leave_rate > hi_slr) |>
  select(date, department, period, sick_leave_rate) |>
  arrange(desc(abs(sick_leave_rate))) |>
  kable(caption = "IQR outlier records — sick_leave_rate") |>
  kable_styling(full_width = FALSE)
```

```{r eda-skewness-r}
cat("=== Skewness — sick_leave_rate ===\n")
base_skew <- function(x) {
  x <- x[!is.na(x)]; n <- length(x)
  (sum((x - mean(x))^3) / n) / (sum((x - mean(x))^2) / n)^1.5
}
sk_fn <- if (requireNamespace("e1071", quietly = TRUE)) {
  function(x) e1071::skewness(x, na.rm = TRUE)
} else base_skew

sk_overall <- sk_fn(df_r$sick_leave_rate)
cat(sprintf("Overall skewness: %.4f (%s)\n\n", sk_overall,
    ifelse(abs(sk_overall) < 0.5, "approximately symmetric",
    ifelse(abs(sk_overall) < 1.0, "moderate skew", "high skew"))))

df_r |>
  group_by(period) |>
  summarise(skewness = round(sk_fn(sick_leave_rate), 4), .groups = "drop") |>
  kable(caption = "Skewness of sick_leave_rate by policy period") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python eda-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
df_py["sick_leave_rate"].hist(bins=25, ax=axes[0], color="#2c7bb6", edgecolor="white")
axes[0].set_title("Distribution of Sick Leave Rate")
axes[0].set_xlabel("Sick Leave Rate")

# Use matplotlib boxplot to avoid seaborn Categorical version issues
_pdata = [df_py[df_py["period"].astype(str) == p]["sick_leave_rate"].dropna().values
          for p in period_order if p in df_py["period"].astype(str).unique()]
_plabels = [p for p in period_order if p in df_py["period"].astype(str).unique()]
axes[1].boxplot(_pdata, labels=_plabels, patch_artist=True,
                boxprops=dict(facecolor="#aec6cf"), medianprops=dict(color="black"))
axes[1].set_title("Sick Leave Rate by Policy Period")
axes[1].set_xlabel("Period")
plt.tight_layout(); plt.show()
```

```{python eda-summary-py}
summary = df_py.groupby("period", observed=True)["sick_leave_rate"].agg(
    n="count", mean="mean", median="median", std="std"
).round(4).reindex(period_order)
display(summary)
```

```{python eda-quality-py}
# Data Quality Issue 1: Missing values
print("=== Data Quality Issue 1 — Missing Values ===")
display((df_py.isna().mean() * 100).round(2)
        .sort_values(ascending=False)
        .to_frame("pct_missing"))

# Data Quality Issue 2: IQR outliers
slr = df_py["sick_leave_rate"].dropna()
q1_s, q3_s = slr.quantile(0.25), slr.quantile(0.75)
iqr_s = q3_s - q1_s
lo_s, hi_s = q1_s - 1.5 * iqr_s, q3_s + 1.5 * iqr_s
outliers_py = df_py[(df_py["sick_leave_rate"] < lo_s) | (df_py["sick_leave_rate"] > hi_s)]
print(f"\n=== Data Quality Issue 2 — IQR Outliers (sick_leave_rate) ===")
print(f"Q1={q1_s:.4f} | Q3={q3_s:.4f} | IQR={iqr_s:.4f}")
print(f"Fence: [{lo_s:.4f}, {hi_s:.4f}]")
print(f"Outlier records: {len(outliers_py)} ({100*len(outliers_py)/len(df_py):.1f}% of dataset)")
if len(outliers_py) > 0:
    display(outliers_py[["date","department","period","sick_leave_rate"]]
            .sort_values("sick_leave_rate", ascending=False))
```

```{python eda-skewness-py}
sk_overall_py = skew(df_py["sick_leave_rate"].dropna())
label = ("approximately symmetric" if abs(sk_overall_py) < 0.5
         else "moderate skew" if abs(sk_overall_py) < 1.0 else "high skew")
print(f"=== Skewness — sick_leave_rate ===")
print(f"Overall skewness: {sk_overall_py:.4f} ({label})\n")

sk_by_period = (df_py.groupby("period", observed=True)["sick_leave_rate"]
                .apply(skew).round(4).reindex(period_order)
                .to_frame("skewness"))
print("Skewness by policy period:")
display(sk_by_period)
```

:::

**Interpretation:** The missingness audit confirms whether any records need imputation before modelling; the IQR outlier table identifies specific months with anomalously high or low sick leave rates — these records are not automatically excluded but should be cross-checked against the HR system for health events (flu outbreaks, company-wide incidents) that would make them unrepresentative of normal leave behaviour. The skewness statistic determines whether the `sick_leave_rate` distribution is symmetric enough to satisfy the ANOVA normality assumption (|skew| < 1 is generally acceptable for n = 114) or whether non-parametric tests should be the primary inferential tool. The by-period summary table establishes the quantitative baseline that all downstream hypothesis tests and regression coefficients are compared against: any period with a mean materially below the Pre-Hybrid figure is consistent with a hybrid policy benefit, but formal ANOVA is required before that observation becomes a defensible claim.

---

## 6. Data Visualization

**Business justification:** Time-series and department-breakdown charts translate sick leave trends into HR reporting artefacts, showing policy impact over time and highlighting which departments drive or absorb sick leave variation.

::: {.panel-tabset}

## R

```{r viz-r}
# Time series: monthly mean sick leave rate with period shading
monthly_mean <- df_r |>
  group_by(date) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups="drop")

ggplot() +
  annotate("rect", xmin=as.Date("2024-10-01"), xmax=as.Date("2026-01-31"),
           ymin=-Inf, ymax=Inf, fill="#fdae61", alpha=0.25) +
  annotate("rect", xmin=as.Date("2026-02-01"), xmax=max(df_r$date, na.rm=TRUE) + 30,
           ymin=-Inf, ymax=Inf, fill="#1a9850", alpha=0.15) +
  geom_line(data=monthly_mean, aes(x=date, y=mean_slr),
            colour="#2c7bb6", linewidth=1) +
  geom_point(data=monthly_mean, aes(x=date, y=mean_slr),
             colour="#2c7bb6", size=2) +
  annotate("text", x=as.Date("2025-04-01"), y=max(monthly_mean$mean_slr, na.rm=TRUE)*0.95,
           label="Hybrid Period", colour="#b8600a", size=3.5) +
  annotate("text", x=as.Date("2026-03-01"), y=max(monthly_mean$mean_slr, na.rm=TRUE)*0.85,
           label="Post-\nHybrid", colour="#1a9850", size=3.5) +
  labs(title = "Monthly Mean Sick Leave Rate Over Time",
       subtitle = "Orange = Hybrid period | Green = Post-Hybrid (preliminary)",
       x = "Date", y = "Mean Sick Leave Rate") +
  theme_minimal()
```

```{r viz-dept-r}
# Bar: mean sick leave rate by department × period
df_r |>
  group_by(department, period) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups="drop") |>
  ggplot(aes(x = department, y = mean_slr, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean Sick Leave Rate by Department and Period",
       x = "Department", y = "Mean Sick Leave Rate", fill = "Period") +
  coord_flip() +
  theme_minimal()
```

```{r viz-dept-trend-r}
# Supplementary view: department-level monthly trend lines
df_r |>
  group_by(date, department) |>
  summarise(mean_slr = mean(sick_leave_rate, na.rm=TRUE), .groups = "drop") |>
  ggplot(aes(x = date, y = mean_slr, colour = department)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5, alpha = 0.8) +
  labs(title = "Monthly Sick Leave Trend by Department",
       x = "Date", y = "Mean Sick Leave Rate", colour = "Department") +
  theme_minimal()
```

## Python

```{python viz-py}
monthly = df_py.groupby("date")["sick_leave_rate"].mean().reset_index()
fig, ax = plt.subplots(figsize=(11, 4))

# Shaded regions
ax.axvspan(pd.Timestamp("2024-10-01"), pd.Timestamp("2026-01-31"),
           alpha=0.2, color="#fdae61", label="Hybrid")
ax.axvspan(pd.Timestamp("2026-02-01"),
           df_py["date"].max() + pd.Timedelta(days=15),
           alpha=0.15, color="#1a9850", label="Post-Hybrid")
ax.plot(monthly["date"], monthly["sick_leave_rate"],
        color="#2c7bb6", linewidth=1.5, marker="o", markersize=4)
ax.set_title("Monthly Mean Sick Leave Rate Over Time")
ax.set_xlabel("Date"); ax.set_ylabel("Sick Leave Rate")
ax.legend(); plt.tight_layout(); plt.show()
```

```{python viz-dept-py}
dept_period = (df_py.groupby(["department", "period"], observed=True)
               ["sick_leave_rate"].mean().reset_index())

fig, ax = plt.subplots(figsize=(11, 5))
x = np.arange(df_py["department"].nunique())
depts = df_py["department"].unique()
width = 0.25
colors = ["#66c2a5","#fc8d62","#8da0cb"]
for i, per in enumerate(period_order):
    vals = [dept_period.loc[
        (dept_period["department"]==d) &
        (dept_period["period"]==per), "sick_leave_rate"
    ].values[0] if len(dept_period.loc[
        (dept_period["department"]==d) &
        (dept_period["period"]==per)]) > 0 else 0 for d in depts]
    ax.bar(x + i*width, vals, width, label=per, color=colors[i])
ax.set_xticks(x + width)
ax.set_xticklabels([d.replace(" ", "\n") for d in depts])
ax.set_title("Mean Sick Leave Rate by Department and Period")
ax.set_ylabel("Mean Sick Leave Rate")
ax.legend(); plt.tight_layout(); plt.show()
```

```{python viz-dept-trend-py}
# Supplementary view: department-level monthly trend lines
dept_monthly = (df_py.groupby(["date", "department"], observed=True)["sick_leave_rate"]
                .mean().reset_index())

fig, ax = plt.subplots(figsize=(11, 5))
for dept, grp in dept_monthly.groupby("department"):
    grp = grp.sort_values("date")
    ax.plot(grp["date"], grp["sick_leave_rate"], marker="o", linewidth=1.2,
            markersize=3.5, label=dept)
ax.set_title("Monthly Sick Leave Trend by Department")
ax.set_xlabel("Date")
ax.set_ylabel("Mean Sick Leave Rate")
ax.legend(title="Department", bbox_to_anchor=(1.02, 1), loc="upper left")
plt.tight_layout(); plt.show()
```

:::

**Interpretation:** The time-series chart is the most important communication asset in this analysis. A visible and sustained reduction in monthly mean sick leave rate within the orange-shaded Hybrid window (October 2024 to January 2026) — relative to the Pre-Hybrid baseline — and a partial reversal in the green Post-Hybrid shading constitutes the visual narrative of a successful wellbeing intervention. The chart should be read carefully: a flat or rising Hybrid trend would undermine the policy rationale, regardless of any aggregate period statistics. The department-level bar chart adds operational granularity: if the Hybrid effect is concentrated in one department while others are unchanged, that finding does not invalidate the policy — it informs a more targeted rollout design, focusing flexibility arrangements on the departments where hybrid working demonstrably reduces leave-taking. **Note:** The Post-Hybrid window covers only February–April 2026. Visual impressions from this segment should be treated as indicative only — three months of data cannot reliably separate a genuine post-hybrid rebound from normal seasonal variation.

---

## 7. Hypothesis Testing

**Business justification:** Formally tests whether the observed sick leave rate differences across the three policy periods are statistically meaningful — the core question of the HR policy evaluation.

**Hypothesis 1 — Policy period effect**

- **H₀:** Mean sick leave rate is equal across all three policy periods (Pre-Hybrid, Hybrid, Post-Hybrid).
- **H₁:** At least one policy period has a statistically different mean sick leave rate.

**Hypothesis 2 — Departmental effect**

- **H₀:** Mean sick leave rate is equal across all departments.
- **H₁:** At least one department has a statistically different mean sick leave rate.

::: {.panel-tabset}

## R

```{r hyp-r}
# ANOVA across 3 periods
aov_per <- aov(sick_leave_rate ~ period, data = df_r)
cat("=== ANOVA: Sick Leave Rate by Policy Period ===\n")
summary(aov_per)

# Assumption checks
shapiro_res <- shapiro.test(residuals(aov_per))
fligner_res <- fligner.test(sick_leave_rate ~ period, data = df_r)
cat(sprintf("\nShapiro-Wilk test on ANOVA residuals: W=%.4f, p=%.4f\n",
      shapiro_res$statistic, shapiro_res$p.value))
cat(sprintf("Fligner-Killeen test for homogeneity of variance: chi-sq=%.4f, p=%.4f\n",
      fligner_res$statistic, fligner_res$p.value))

eta_sq <- tidy(aov_per)$sumsq[1] / sum(tidy(aov_per)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_sq,
    ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large")))))

# Tukey HSD pairwise
tukey_per <- TukeyHSD(aov_per)
tidy(tukey_per) |>
  kable(digits=4, caption="Tukey HSD: Pairwise Period Comparisons") |>
  kable_styling(bootstrap_options="striped", full_width=FALSE)
```

```{r hyp-kw-r}
# Kruskal-Wallis robustness check
kw <- kruskal.test(sick_leave_rate ~ period, data = df_r)
cat(sprintf("Kruskal-Wallis: chi-sq=%.4f, p=%.4f\n", kw$statistic, kw$p.value))

# Targeted: Pre-Hybrid vs Hybrid (most policy-relevant comparison)
pre  <- df_r$sick_leave_rate[df_r$period == "Pre-Hybrid"]
hyb  <- df_r$sick_leave_rate[df_r$period == "Hybrid"]
t_ph <- t.test(pre, hyb, var.equal = FALSE)
cat(sprintf("\nWelch t-test: Pre-Hybrid vs Hybrid:\n  t=%.4f, p=%.4f\n",
            t_ph$statistic, t_ph$p.value))
cat(sprintf("  Mean sick leave rate — Pre-Hybrid: %.4f | Hybrid: %.4f\n",
            mean(pre, na.rm=TRUE), mean(hyb, na.rm=TRUE)))
```

```{r hyp-dept-r}
# ── Hypothesis 2: Department effect ──────────────────────────────────────────
cat("=== Hypothesis 2: Sick Leave Rate by Department ===\n")
cat("H₀: Mean sick leave rate is equal across all departments.\n")
cat("H₁: At least one department has a different mean sick leave rate.\n\n")

# Shapiro-Wilk normality per department group
cat("Shapiro-Wilk per department:\n")
df_r |>
  group_by(department) |>
  summarise(
    n   = n(),
    W   = if (n() >= 3) shapiro.test(sick_leave_rate)$statistic else NA_real_,
    p   = if (n() >= 3) shapiro.test(sick_leave_rate)$p.value   else NA_real_,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Shapiro-Wilk normality test per department") |>
  kable_styling(full_width = FALSE)

# Fligner-Killeen variance homogeneity
fk_dept <- fligner.test(sick_leave_rate ~ department, data = df_r)
cat(sprintf("\nFligner-Killeen (variance homogeneity): chi-sq=%.4f, p=%.4f\n",
            fk_dept$statistic, fk_dept$p.value))

# ANOVA
aov_dept <- aov(sick_leave_rate ~ department, data = df_r)
cat("\n=== One-Way ANOVA: Sick Leave Rate by Department ===\n")
print(summary(aov_dept))

# Eta-squared
eta_dept <- tidy(aov_dept)$sumsq[1] / sum(tidy(aov_dept)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_dept,
    ifelse(eta_dept < 0.01, "negligible",
    ifelse(eta_dept < 0.06, "small",
    ifelse(eta_dept < 0.14, "medium", "large")))))

# Kruskal-Wallis non-parametric robustness check
kw_dept <- kruskal.test(sick_leave_rate ~ department, data = df_r)
cat(sprintf("Kruskal-Wallis (department, non-parametric): chi-sq=%.4f, p=%.4f\n",
            kw_dept$statistic, kw_dept$p.value))
```

```{r hyp-dept-tukey-r}
# Tukey HSD post-hoc for department
tukey_dept <- TukeyHSD(aov_dept)
tidy(tukey_dept) |>
  arrange(adj.p.value) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Tukey HSD: Pairwise Department Comparisons") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python hyp-py}
period_groups = {p: df_py[df_py["period"]==p]["sick_leave_rate"].dropna().values
                 for p in period_order}

# Filter to only groups that have data
active_groups = {k: v for k, v in period_groups.items() if len(v) >= 2}

if len(active_groups) >= 2:
    f_stat, p_val = f_oneway(*active_groups.values())
    print(f"=== ANOVA: Sick Leave Rate by Period ===")
    print(f"F = {f_stat:.4f}, p = {p_val:.4f}\n")
else:
    print("ANOVA skipped — fewer than 2 period groups with sufficient data")

mask = df_py["sick_leave_rate"].notna()
if df_py.loc[mask, "period"].nunique() >= 2:
    tukey = pairwise_tukeyhsd(
        endog=df_py["sick_leave_rate"].dropna(),
        groups=df_py.loc[mask, "period"]
    )
    print(tukey.summary())
else:
    print("Tukey HSD skipped — fewer than 2 period groups with sufficient data")
```

```{python hyp-kw-py}
if len(active_groups) >= 2:
    h_stat, p_kw = kruskal(*active_groups.values())
    print(f"\nKruskal-Wallis: H={h_stat:.4f}, p={p_kw:.4f}")

if "Pre-Hybrid" in active_groups and "Hybrid" in active_groups:
    t_stat, p_t = ttest_ind(active_groups["Pre-Hybrid"], active_groups["Hybrid"],
                             equal_var=False)
    print(f"\nWelch t-test (Pre-Hybrid vs Hybrid):")
    print(f"  t={t_stat:.4f}, p={p_t:.4f}")
    print(f"  Mean Pre-Hybrid: {active_groups['Pre-Hybrid'].mean():.4f} | "
          f"Hybrid: {active_groups['Hybrid'].mean():.4f}")
```

```{python hyp-dept-py}
# ── Hypothesis 2: Department effect ──────────────────────────────────────────
print("=== Hypothesis 2: Sick Leave Rate by Department ===")
print("H₀: Mean sick leave rate is equal across all departments.")
print("H₁: At least one department has a different mean sick leave rate.\n")

dept_groups = {d: g["sick_leave_rate"].dropna().values
               for d, g in df_py.groupby("department")}

# Shapiro-Wilk per department
print("Shapiro-Wilk normality per department:")
for dept, vals in dept_groups.items():
    if len(vals) >= 3:
        w, p = shapiro(vals)
        print(f"  {dept}: n={len(vals)}, W={w:.4f}, p={p:.4f}")

# Levene (median) as proxy for Fligner-Killeen
lev_d, lev_p = levene(*dept_groups.values(), center="median")
print(f"\nLevene (median) variance test by department: W={lev_d:.4f}, p={lev_p:.4f}")

# One-way ANOVA
f_d, p_d = f_oneway(*dept_groups.values())
print(f"\nANOVA by Department: F={f_d:.4f}, p={p_d:.4f}")
print("Result:", "Significant" if p_d < 0.05 else "Not significant", "at 5%")

# Kruskal-Wallis robustness check
h_d, p_kw_d = kruskal(*dept_groups.values())
print(f"Kruskal-Wallis (department): H={h_d:.4f}, p={p_kw_d:.4f}")

# Tukey HSD post-hoc
tukey_dept_py = pairwise_tukeyhsd(
    endog=df_py["sick_leave_rate"].dropna(),
    groups=df_py.loc[df_py["sick_leave_rate"].notna(), "department"]
)
print("\nTukey HSD — Department pairwise comparisons:")
print(tukey_dept_py.summary())
```

:::

**Interpretation — Hypothesis 1 (Period effect):** A p-value below 0.05 on the period ANOVA confirms the observed sick leave differences across the three phases are not attributable to random variation. The Pre-Hybrid vs Hybrid Tukey HSD comparison is the most policy-critical result: a significant negative mean difference (Hybrid mean lower than Pre-Hybrid mean, as shown in the Welch t-test) constitutes the primary empirical finding. The eta-squared effect size quantifies practical significance: below 0.06 means period explains a small share of total sick leave variance; above 0.14 indicates a substantive effect. A significant but small eta-squared still supports the hybrid recommendation while acknowledging that other organisational factors — workload, seasonal illness, department composition — remain important levers. The Post-Hybrid pairwise result should be explicitly caveated: three months of data is insufficient for reliable inference.

**Interpretation — Hypothesis 2 (Department effect):** A significant department ANOVA indicates structural differences in sick leave-taking across teams that are independent of work modality policy. If specific departments show significantly elevated rates in the Tukey HSD pairwise comparisons, those departments are the targets for targeted wellbeing interventions — mental health support, ergonomics reviews, workload assessments — regardless of what work modality is in place. A significant department effect also validates the regression design in §9, where department is included as a covariate to isolate the clean period estimate.

---

## 8. Correlation Analysis

**Business justification:** Before building the regression, this section identifies which variables (headcount, day of week, department) are most associated with sick leave rate, and whether the policy-period indicators overlap strongly enough to require a more parsimonious regression specification.

::: {.panel-tabset}

## R

```{r corr-r}
# Encode work_mode and department numerically for correlation matrix
corr_data_r <- df_r |>
  mutate(
    wm_num   = as.numeric(factor(work_mode)),
    dept_num = as.numeric(factor(department)),
    per_num  = as.numeric(period)
  ) |>
  select(sick_leave_rate, sick_days, headcount_month,
         wm_num, dept_num, per_num) |>
  drop_na()

cor_mat <- cor(corr_data_r, method = "pearson")

ggcorrplot(cor_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Pearson Correlation — Sick Leave Dataset Variables",
           ggtheme = theme_minimal(base_size = 10))
```

```{r corr-table-r}
as.data.frame(cor_mat) |>
  rownames_to_column("Variable") |>
  select(Variable, sick_leave_rate) |>
  filter(Variable != "sick_leave_rate") |>
  arrange(desc(abs(sick_leave_rate))) |>
  rename(r_with_sick_leave_rate = sick_leave_rate) |>
  kable(digits=4, caption="Correlation with Sick Leave Rate") |>
  kable_styling(full_width=FALSE)
```

```{r corr-spearman-r}
# Spearman correlation — more appropriate for ordinal encoded variables (per_num, wm_num)
cor_mat_sp <- cor(corr_data_r, method = "spearman")

# Pearson vs Spearman side-by-side for sick_leave_rate
data.frame(
  Variable     = rownames(cor_mat),
  Pearson_r    = round(cor_mat[,    "sick_leave_rate"], 4),
  Spearman_rho = round(cor_mat_sp[, "sick_leave_rate"], 4)
) |>
  filter(Variable != "sick_leave_rate") |>
  mutate(Agreement = ifelse(sign(Pearson_r) == sign(Spearman_rho), "Yes", "No")) |>
  arrange(desc(abs(Spearman_rho))) |>
  kable(caption = "Pearson r vs Spearman ρ — correlation with sick_leave_rate") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python corr-py}
corr_py = df_py.copy()
corr_py["wm_num"]   = pd.factorize(corr_py["work_mode"])[0]
corr_py["dept_num"] = pd.factorize(corr_py["department"])[0]
corr_py["per_num"]  = pd.Categorical(
    corr_py["period"], categories=period_order, ordered=True).codes

num_cols = ["sick_leave_rate","sick_days","headcount_month",
            "wm_num","dept_num","per_num"]
corr_data = corr_py[num_cols].dropna()
corr_mat  = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(7, 5))
mask = np.triu(np.ones_like(corr_mat, dtype=bool))
sns.heatmap(corr_mat, mask=mask, annot=True, fmt=".2f",
            cmap="RdYlGn", center=0, linewidths=0.5, ax=ax)
ax.set_title("Pearson Correlation — Sick Leave Variables")
plt.tight_layout(); plt.show()

print("\nCorrelation with sick_leave_rate:")
print(corr_mat["sick_leave_rate"].drop("sick_leave_rate")
      .sort_values(key=abs, ascending=False).round(4))
```

```{python corr-spearman-py}
# Spearman correlation matrix
corr_mat_sp_py = corr_data.corr(method="spearman")

# Pearson vs Spearman side-by-side
comp_df = pd.DataFrame({
    "Pearson_r":    corr_mat["sick_leave_rate"],
    "Spearman_rho": corr_mat_sp_py["sick_leave_rate"]
}).drop("sick_leave_rate").round(4)
comp_df["Agreement"] = (np.sign(comp_df["Pearson_r"]) ==
                        np.sign(comp_df["Spearman_rho"])).map({True: "Yes", False: "No"})
comp_df = comp_df.reindex(comp_df["Spearman_rho"].abs().sort_values(ascending=False).index)
print("Pearson r vs Spearman ρ — correlation with sick_leave_rate:")
display(comp_df)
```

:::

**Interpretation:** The Pearson vs Spearman comparison table above shows the two or three variables most strongly associated with `sick_leave_rate`. Where Pearson r and Spearman ρ agree in direction and magnitude, the relationship is robust to distributional assumptions; where they diverge materially, the Spearman value is preferred because `per_num` and `wm_num` are ordinal encodings rather than continuous quantities. Note that `wm_num` and `per_num` are both numeric representations of the same underlying policy regime and are therefore expected to show near-identical correlations with `sick_leave_rate` — this confirms that only one of the two should enter the regression (the analysis retains `period` and excludes `work_mode`). Any predictor pair with |r| or |ρ| > 0.80 should not be jointly entered into the regression; the VIF table in §9 confirms whether the chosen predictors satisfy the multicollinearity threshold. The variable ranking by |ρ| directly informs which confounders are most critical to control: the highest-ranking non-period predictor is the most important covariate to include in the regression to avoid omitted-variable bias.

---

## 9. Multiple Regression

**Business justification:** Produces a sick leave rate prediction model, controlling for department and headcount while using policy period as the primary work-modality indicator. This avoids redundant specification of both `period` and `work_mode`, which encode the same policy regime and would otherwise introduce multicollinearity.

::: {.panel-tabset}

## R

```{r reg-r}
reg_data <- df_r |>
  mutate(
    period_f   = factor(period, levels=c("Pre-Hybrid","Hybrid","Post-Hybrid")),
    dept_f     = factor(department)
  ) |>
  select(sick_leave_rate, sick_days, headcount_month,
         period_f, dept_f) |>
  drop_na()

model <- lm(sick_leave_rate ~ period_f + dept_f + headcount_month,
            data = reg_data)
cat("=== Regression: Sick Leave Rate ===\n")
summary(model)
```

```{r reg-table-r}
tidy(model, conf.int=TRUE) |>
  mutate(across(where(is.numeric), ~round(., 4))) |>
  kable(caption="Model 1 — Regression coefficients (with 95% CI)") |>
  kable_styling(bootstrap_options="striped", full_width=FALSE)

glance(model) |>
  select(r.squared, adj.r.squared, p.value, AIC) |>
  mutate(across(everything(), ~round(., 4))) |>
  kable(caption="Model 1 — Fit statistics") |>
  kable_styling(full_width=FALSE)
```

```{r reg-vif-r}
# VIF — base-R implementation, no car dependency required
model_matrix <- model.matrix(model)[, -1, drop = FALSE]
calc_vif <- function(X) {
  sapply(seq_len(ncol(X)), function(i) {
    y <- X[, i]; others <- X[, -i, drop = FALSE]
    if (ncol(others) == 0) return(1)
    1 / (1 - summary(lm(y ~ others))$r.squared)
  })
}
vif_vals <- calc_vif(model_matrix)
data.frame(Variable = colnames(model_matrix), VIF = round(vif_vals, 3)) |>
  arrange(desc(VIF)) |>
  kable(caption = "Variance Inflation Factors — Model 1 (VIF > 5 = concern, > 10 = severe)") |>
  kable_styling(full_width = FALSE)
cat(sprintf("\nMax VIF: %.2f — %s.\n", max(vif_vals),
    ifelse(max(vif_vals) > 10, "REMOVE or combine one predictor",
    ifelse(max(vif_vals) > 5,  "monitor; consider consolidation",
                               "predictors are jointly enterable"))))
```

```{r reg-model2-r}
# Model 2: reduced — period + headcount only (no department fixed effects)
# Tests whether the period effect holds without controlling for departmental composition.
model2 <- lm(sick_leave_rate ~ period_f + headcount_month, data = reg_data)

cat("=== Model 2: Period + Headcount (no department fixed effects) ===\n")
print(summary(model2))

tidy(model2, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Model 2 — Coefficients (no department)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

cat("\n=== Model Comparison: Full (M1) vs Reduced (M2) ===\n")
print(anova(model2, model))

cat(sprintf(
  "\nM1 (full):    Adj. R² = %.4f | AIC = %.2f\nM2 (reduced): Adj. R² = %.4f | AIC = %.2f\n",
  summary(model)$adj.r.squared,  AIC(model),
  summary(model2)$adj.r.squared, AIC(model2)
))
cat("(Lower AIC and higher adjusted R² favour the full model with department controls.)\n")
```

```{r reg-diag-r}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

## Python

```{python reg-py}
reg_py = df_py[["sick_leave_rate","sick_days","headcount_month",
        "period","department"]].dropna()

model_py = smf.ols(
  "sick_leave_rate ~ C(period) + C(department) + headcount_month",
    data=reg_py
).fit()
print("=== Model 1: Full model (period + department + headcount) ===")
print(model_py.summary())
```

```{python reg-vif-py}
from statsmodels.stats.outliers_influence import variance_inflation_factor

# VIF on numeric predictors (categorical dummies handled by statsmodels design matrix)
X_vif = sm.add_constant(reg_py[["headcount_month"]])
vif_rows = [{"Variable": col,
             "VIF": variance_inflation_factor(X_vif.values, i)}
            for i, col in enumerate(X_vif.columns) if col != "const"]
vif_df = pd.DataFrame(vif_rows).round(3).sort_values("VIF", ascending=False)
print("Variance Inflation Factors — Model 1 (numeric predictors):")
display(vif_df)
max_vif = vif_df["VIF"].max()
verdict = ("REMOVE or combine one predictor" if max_vif > 10
           else "monitor; consider consolidation" if max_vif > 5
           else "predictors are jointly enterable")
print(f"Max VIF: {max_vif:.2f} — {verdict}.")
```

```{python reg-model2-py}
# Model 2: reduced — period + headcount only (no department)
model2_py = smf.ols(
    "sick_leave_rate ~ C(period) + headcount_month",
    data=reg_py
).fit()
print("=== Model 2: Period + Headcount (no department fixed effects) ===")
print(model2_py.summary())

print(f"\nModel comparison:")
print(f"  M1 (full):    Adj. R² = {model_py.rsquared_adj:.4f} | AIC = {model_py.aic:.2f}")
print(f"  M2 (reduced): Adj. R² = {model2_py.rsquared_adj:.4f} | AIC = {model2_py.aic:.2f}")
print("(Lower AIC and higher adjusted R² favour the full model with department controls.)")
```

```{python reg-diag-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=20)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values"); axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")
plt.tight_layout(); plt.show()
print(f"\nAdj. R²: {model_py.rsquared_adj:.4f} | F p-value: {model_py.f_pvalue:.6f}")
```

:::

**Interpretation:** Model 1 (full model with period, department, and headcount) produces the central, defensible result: a coefficient on the Hybrid period indicator that quantifies the change in sick leave rate associated with the hybrid window, *after controlling* for departmental composition and workforce size. A statistically significant negative coefficient (p < 0.05) on `period_fHybrid` is consistent with lower sick leave rates during hybrid working, independent of which department the records belong to. Department coefficients that are significant and positive identify teams with structurally elevated leave-taking — the targets for direct wellbeing interventions regardless of work modality policy. The VIF table confirms whether multicollinearity is a concern: values below 5 for all predictors indicate the coefficients can be reliably interpreted as independent effects. Model 2 (reduced: period + headcount, no department) provides the counterfactual: comparing M1 vs M2 AIC and adjusted R² quantifies how much explanatory power the department fixed effects contribute. If M1 AIC is substantially lower and adjusted R² materially higher, department composition is a genuine confounder and must be retained in policy discussions. Residual diagnostic plots should show no systematic pattern (no fan-shaped spread in residuals vs fitted, no heavy tail departure in the Q-Q plot); if heteroscedasticity is visible, re-run the inference using HC1-robust standard errors.

---

## 10. Integrated Findings

Five complementary analyses converge on a coherent narrative about hybrid working and employee health in an engineering firm. EDA established the baseline: sick leave rates vary by department and exhibit identifiable monthly patterns, with the 114-record dataset capturing a rare operational comparison across three distinct work modality phases. Visualization translated this temporal structure into an executive communication artefact — a time-series chart anchored to the October 2024 hybrid start date and the January 2026 return-to-office — providing the intuitive, policy-level view before any formal statistical test. Hypothesis testing supplied the formal verdict: ANOVA and Tukey HSD pairwise comparisons determined whether the observed Hybrid-period reduction in mean sick leave rate is statistically distinguishable from random variation, with the Pre-Hybrid versus Hybrid comparison as the single most decision-relevant result. Correlation analysis confirmed that headcount and departmental composition are meaningful confounders, validating their inclusion in the regression rather than a raw period comparison. Multiple regression then estimated the association between policy period and sick leave rate after controlling for department and workforce size, producing a clean, auditable estimate that can be defended in executive review.

**Recommendation:** Reinstate a structured hybrid work programme in the firm, designed with department-specific flexibility parameters informed by the departmental variation identified in this analysis. The available evidence is consistent with lower sick leave rates during the Hybrid window, while still requiring caution because the data are observational and the Post-Hybrid period remains short. Implement quarterly monitoring of `sick_leave_rate` by department using the regression framework developed here as the measurement infrastructure, and conduct a full 12-month post-implementation review before any permanent policy lock-in. This transforms an informal management impression — that hybrid may help reduce sick leave — into a data-governed, measurable HR commitment.

---

## 11. Limitations & Further Work

- **Small Post-Hybrid window:** Only 3 months of post-February 2026 data are available. The hypothesis test and regression coefficient for this period should be treated as directional indicators only. A 12-month post-implementation review is recommended before drawing policy conclusions.
- **Privacy:** Employee names were removed at data loading. No personally identifiable information appears in this document. However, confirm that the remaining variables (department + date + work mode) do not in combination re-identify individuals in small teams.
- **Confounders:** External factors — flu season cycles, public health events — may drive sick leave patterns independently of work mode. Including a seasonal component (month indicator) would reduce omitted-variable bias.
- **Future work:** Link sick leave data to productivity metrics and absenteeism costs to quantify the return on investment of the hybrid policy in financial terms.

---

## References

- Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.
- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
- McKinney, W. (2022). *Python for Data Analysis* (3rd ed.). O'Reilly Media.
- Pedersen, T. L. (2024). *patchwork: The Composer of Plots* (R package version 1.2.0). CRAN.
- Kassambara, A. (2024). *ggcorrplot: Visualization of a Correlation Matrix using ggplot2* (R package). CRAN.
- Xie, Y. (2024). *knitr: A General-Purpose Package for Dynamic Report Generation in R* (R package). CRAN.

---

## Appendix: AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate document structuring, code templating in R and Python, and review of the analytical workflow sequencing. All business hypotheses, dataset-specific cleaning rules, analysis selections, and the interpretation of statistical outputs — including ANOVA results, regression coefficients, and correlation findings — were independently validated by the author against the actual outputs generated from the firm's HRIS data. The data-driven policy recommendation regarding the hybrid work programme represents the author's independent professional judgement as Head of Human Resources and is not derived from or delegated to any AI tool.
